---
title: "For Live Session 2"
author: "Eric Graham"
date: '`r Sys.Date()`'
output: powerpoint_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tswge)
```

```{r, include=FALSE}
cpi_data = read.csv("CPIAUCSL.csv", header = TRUE)
str(cpi_data)
x = cpi_data$CPIAUCSL

cat("x1 =", x[1], "\n")
cat("x2 =", x[2], "\n") 
cat("x3 =", x[3], "\n")
cat("Total observations:", length(x), "\n")

plotts.wge(x)
acf(x[1:250], plot = TRUE)
```

# Spectral Density and Frequencies

## Spectral Density 

The spectral density plot shows a peak at frequency 0.0, indicating that there is no periodic cycle.

## Frequencies

No distinct periodic frequencies detected. The peak near zero confirms trending behavior that we would expect from non-stationary data.

```{r}
parzen.wge(x)
```

# Relevance to the Series

These results are in alignment with what one would expect for CPI as an inflationary indicator: there are persistent upward price trends rather than seasonal cycles, confirming the trend in the time series. I hope there's a transformation for this because I might have picked the wrong dataset!

# Key Takeaways

* Time series can be analyzed by converting them into sin and cosine expressions using Fourier expansion to identify underlying frequencies
* The spectral density estimates which frequencies are present in data, with peaks at zero indicating trending behavior.
* Raw spectral density estimates are noisy and require smoothing (windowing) to reveal meaningful patterns, using a truncation parameter M.