---
title: "Unit 5 FLS"
author: "Eric Graham"
date: "`r Sys.Date()`"
output: 
  powerpoint_presentation:
    slide_level: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tswge)
```

# 1: Inflation Data

```{r}
cpi = read.csv("CPIAUCSL.csv")
plotts.sample.wge(cpi$CPIAUCSL)

mean(cpi$CPIAUCSL)
var(cpi$CPIAUCSL)

acf(cpi$CPIAUCSL)
```

Revisiting the Consumer Price Index data, we can now begin to assess its suitability for an ARIMA model.

As previously noted, the time series is non-stationary: there is clear evidence of an upward trend which violates the constant mean assumption, and the high, slowly-dampening autocorrelations indicate strong wandering behavior. An ARIMA model may be helpful here, as differencing can help us achieve stationarity.

# 2: Factor Table for S=7

```{r}
factor.wge(phi=c(rep(0,6),1))
```

# 3: Categorizing Models

## Model 1

This model has a seasonal factor (1-B^4) and both MA factors are invertible, so this would be an ARUMA with d = 0 and s≠0 (p=0 because no stationary AR component and q = 3 because that's the highest power of the combined MA factors)

```{r}
factor.wge(phi = c(0, 0, 0, 1))

factor.wge(phi = c(-.3))

factor.wge(phi = c(.6, -.8))
```

## Model 2

Both the AR and MA factors have absolute reciprocals < 1, indicating a stationary and invertible ARMA model (p=1, q=2 because of highest powers in the AR and MA components).

```{r}
factor.wge(phi = c(-.3))

factor.wge(phi = c(.6, -.8))
```

## Model 3

The AR6 component is stationary (absolute reciprocals all <1). 

The (1-B)^2 component has two roots on the unit circle (absolute reciprocal = 1) which confirms that d=2.

The (1-B^12) component has seven roots on the unit circle (absolute reciprocal = 1) and is a seasonal factor with s=12.

The MA component has an absolute reciprocal < 1 so it is invertible. 

This is an ARUMA with d=2 and s=12, p=6 (highest power from the AR6) and q=2 (from the MA side).

```{r}
factor.wge(phi = c(.1, .99, -.013, -.2078, -.0888, -.00864))

factor.wge(phi = c(2, -1))

factor.wge(phi = c(rep(0, 11), 1))

factor.wge(phi = c(.6, -.8))
```

# 4: One Year of Wal-Mart Stock Data

```{r}
wmt = read.csv("wmt.csv")
plotts.sample.wge(wmt$Close)

wmt_diff = artrans.wge(wmt$Close, phi.tr = 1)
```

## 

```{r}
plotts.sample.wge(wmt_diff)

aic5.wge(wmt_diff)
```

Because the differenced data appears to be a random walk, I would start with an ARIMA() model.

# 5: Key Takeaways

* Non-stationary models include signal plus noise, ARIMA models, and seasonal ARUMA models.
* ARIMA models require differencing to achieve stationarity.
* ARUMA models combine AR and MA components with differencing and seasonal factors to create flexible models for complex real-world data
* Seasonal models have factors like (1-B^12) for monthly data or (1-B^4) for quarterly data.
* The artrans.wge() function can be used to implement differencing on a time series in R
